Dependently Typed Languages in Statix

### 2022-10-14 paper-evcs/oasics-v2021-sample-article.tex
Work on paper

The implementation of the language is available at \url{https://github.com/JonathanBrouwer/master-thesis/}. The \verb|final-simple| branch contains the implementation that is relevant for this paper.

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type-checking rules for the language. We use some syntax in order to make the rules a bit more compact. 
\begin{enumerate}
	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

%\begin{figure}
%\begin{mathpar}
%\inferrule{ }{ s, Type(), [] \Rightarrow s, Type() }
%
%\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
%\end{mathpar}
%\caption{Rules for beta reducing the Calculus of Constructions}
%\label{fig:beta-head-rules}
%\end{figure}

\section{Extensions}

\subsection{Avoiding variable capture}
One problem with the current implementation is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, the correct type is not possible to express without renaming!

The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names.

\subsection{Inference}

Ideally,

\subsection{Data Types}

>>>>

The implementation of the language is available on github. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
\url{https://github.com/JonathanBrouwer/master-thesis/}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type-checking rules for the language. We use some syntax in order to make the rules a bit more compact. 
\begin{enumerate}
	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

\section{Extensions}
This section will discuss some further ideas that can be explored to build upon what has already been shown. These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.

\paragraph*{Avoiding variable capture}
One problem with the current implementation is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, the correct type is not possible to express without renaming! The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names.

\paragraph*{Term Inference}
In dependently typed language, we can infer the value of types. Ideally, we would like to infer the most general type possible. However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not. We implemented an approximation to inference that can infer most types.

\paragraph*{Inductive Data Types}
Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types. Both of these are also fully possible to express in Statix.

\paragraph*{Semantic Code Completion}
Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It functions works well, only showing completions that are semantically possible.
